The yaq Project: \\ Standardized Software Enabling Flexible Instrumentation

### 2022-11-17 paper.tex
JCW comments

Instrumentation development is a key part of the scientific enterprise.
Novel instruments are typically constructed of many individual components that are both purchased and home-built.
Orchestration software must communicate with each hardware component in the course of a scientific experiment.
This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
The challenge of integrating all of these interfaces is a frustrating piece of the modern instrument development process.
Weeks can be spent just integrating one new component into an existing project.
In small academic labs, these software interfaces are typically created by student researchers without software development experience.
Student researchers rarely focus on software reusability, and a lack of maintenance and documentation can make such software more difficult to use as time goes on.
Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.

@@

Figure \ref{fig:network} diagrams the \yaq{} architecture.
Here, we show three different computers connected via an Ethernet network.
The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the network.
This might represent a complex scientific instrument involving several operator terminals as well as embedded computers.
At the top, a single computer is connected to four hardware peripherals through RS232, TTL, USB, and PCI as indicated by the colored lines.
That same computer is running four separate programs, one for each peripheral.
These small, targeted, programs are managed by the operating system and run in the background.
It is conventional to call such programs ``daemons''\cite{Raymond_1996}.
The middle computer is connected to two additional peripherals, and runs daemons for each.
Besides communicating with the hardware peripheral, each daemon can communicate with other programs, ``clients'',  through the network.
The four client programs shown in Figure \ref{fig:network} can each communicate with all six hardware peripherals shown.
As an example, a client running on the bottom computer could communicate with the RS232 peripheral shown in green via the following path: client $\leftrightarrow$ network switch $\leftrightarrow$ top computer $\leftrightarrow$ daemon $\leftrightarrow$ hardware peripheral.
This powerful architecture can be used on a single computer or used across many networked computers, including fully remote operator interfaces.
This client-server architecture offers similar network capabilities to EPICS and TANGO.
As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders outside of large facilities.

@@

In the middle, the same three hardware peripherals are represented in an interactive, graphical form.
\texttt{yaqc-qtpy}\footnote{\texttt{yaqc-qtpy} available on PyPI at https://pypi.org/project/yaqc-qtpy/} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
The self-describing \yaq{} interface is used to provide graphical elements for the most commonly used messages.
This is an invaluable tool which provides a ``free'' graphical user interface (GUI) to any daemon.

@@

The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a, SundenKyle2022}.
Now, using \yaq{}, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
This allows Wright Group researchers to create polished client interfaces without interrupting  ongoing experiments.
Clients developed by the Wright Group have proved flexible enough to be used on four laser systems, each with different complements of hardware.
Moving forward, the Wright Group will spend less energy developing control software and more energy developing creative spectroscopy experiments.

@@

In a \yaq{} context, new hardware can be incorporated into an instrument through the addition of a new daemon.
The \yaq{} architecture simplifies hardware interface development in several ways.
First, because daemons are separate and portable programs, the development effort can be spread across the community of \yaq{} users.
Often, researchers can download an existing daemon rather than writing a new one.
Second, \yaq{} daemon development can be performed separated from the particulars of any individual client.
Often, this means that initial hardware enablement work can be done on a researchers personal machine before the new hardware peripheral is installed in the instrument.
Third, when developing trait-compliant protocols, it becomes easy to design and fully test your hardware interface.
Traits are unambiguous and well-described, making an obvious target for development.
Tooling exists to verify full trait compliance, for example you can use \texttt{yaqc-qtpy} to provide a graphical program to interact with your hardware immediately.
Finally, as discussed in Section II, \yaq{} provides options to design using remote hardware or unusual interfaces when necessary.

@@

Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
Most interfaces, such as Brooks MFC\footnote{\texttt{yaqd-brooks-mfc-gf} source code available on GitHub at \texttt{https://github.com/yaq-project/yaqd-brooks/}}, have been implemented in fewer than 100 lines of Python code.
Even the most complicated daemons are implemented in about 300 lines.
Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{pyserial, pyusb, pyvisa, pymodbus}.
In our experience, the process of creating a new daemon involves about a day of work after communication with the hardware using Python is well understood.
There are currently 72 daemons in the yaq project supporting at least 66 types of hardware, noting that some daemons support the same hardware and others are software only.
Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
A living list of all daemons and supported hardware can be found on the \yaq{} website.\footnote{See https://yaq.fyi/daemons/ and https://yaq.fyi/hardware/ for living lists of all \yaq{} daemons and supported hardware, respectively.}

@@

\yaq{} is open source software.
Anyone can view, install, edit, and suggest changes to our growing collection of daemons and clients.
Furthermore, anyone can create their own totally-custom client or daemon software separately by following the specified \yaq{} standard.
Thirteen individuals have contributed code to the development of \yaq{}.
Open-source development can be a powerful approach for research communities looking to share software-development and maintenance burdens. \cite{CohenJeremy2021a}
It is our hope that a vibrant open-source community will form around \yaq{}.
While open-source development is not a panacea \cite{NowogrodzkiAnna2019a}, we hope that by maintaining a distributed development strategy with a strict focus on only hardware interfaces the \yaq{} project might prove sustainable.

>>>>

Instrumentation development is a key part of the scientific enterprise.
Novel instruments are typically constructed of many individual components that are both purchased and home-built.
Orchestration software must communicate with each hardware component in the course of a scientific experiment.
Orchestration can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
The challenge of integrating all of these interfaces is a frustrating piece of the modern instrument development process.
Weeks can be spent just integrating one new component into an existing project.
In small academic labs, these software interfaces are typically created by student researchers without software development experience.
Student researchers rarely focus on software reusability, and a lack of maintenance and documentation can make such software more difficult to use as time goes on.
Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.

@@

Figure \ref{fig:network} diagrams the \yaq{} architecture.
Here, we show three different computers connected via an Ethernet network.
The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the networks.
This diagram might represent a complex scientific instrument involving several operator terminals as well as embedded computers.
At the top, a single computer is connected to four hardware peripherals through RS232, TTL, USB, and PCI as indicated by the colored lines.
That same computer is running four separate programs, one for each peripheral.
These small, targeted, programs are managed by the operating system and run in the background.
It is conventional to call such programs ``daemons''\cite{Raymond_1996}.
The middle computer is connected to two additional peripherals, and runs daemons for each.
Besides communicating with the hardware peripheral, each daemon can communicate with other programs, ``clients'',  through the network.
The four client programs shown in Figure \ref{fig:network} can each communicate with all six hardware peripherals shown.
As an example, a client running on the bottom computer could communicate with the RS232 peripheral shown in green via the following path: client $\leftrightarrow$ network switch $\leftrightarrow$ top computer $\leftrightarrow$ daemon $\leftrightarrow$ hardware peripheral.
This powerful architecture can be used on a single computer or used across many networked computers, including fully remote operator interfaces.
This client-server architecture offers similar network capabilities to EPICS and TANGO.
As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders outside of large facilities.

@@

In the middle, the same three hardware peripherals are represented in an interactive, graphical form.
\texttt{yaqc-qtpy}\footnote{\texttt{yaqc-qtpy} available on PyPI at https://pypi.org/project/yaqc-qtpy/} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
The self-describing \yaq{} interface is used to provide graphical elements for the most commonly used messages.
\texttt{yaqc-qtpy} is an invaluable tool which provides a ``free'' graphical user interface (GUI) to any daemon.

@@

The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a, SundenKyle2022}.
Now, using \yaq{}, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
Hardware simulation allows Wright Group researchers to create polished client interfaces without interrupting  ongoing experiments.
Clients developed by the Wright Group have proved flexible enough to be used on four laser systems, each with different complements of hardware.
Moving forward, the Wright Group will spend less energy developing control software and more energy developing creative spectroscopy experiments.

@@

In a \yaq{} context, new hardware can be incorporated into an instrument through the addition of a new daemon.
The \yaq{} architecture simplifies hardware interface development in several ways.
First, because daemons are separate and portable programs, the development effort can be spread across the community of \yaq{} users.
Often, researchers can download an existing daemon rather than writing a new one.
Second, \yaq{} daemon development can be performed separately from the particulars of any individual client.
Often, this separation allows initial hardware enablement work to be done on a researchers personal machine before the new hardware peripheral is installed in the instrument.
Third, when developing trait-compliant protocols, it becomes easy to design and fully test your hardware interface.
Traits are unambiguous and well-described, making an obvious target for development.
Tooling exists to verify full trait compliance, for example you can use \texttt{yaqc-qtpy} to provide a graphical program to interact with your hardware immediately.
Finally, as discussed in Section II, \yaq{} provides options to design using remote hardware or unusual interfaces when necessary.

@@

Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
Most interfaces, such as Brooks MFC\footnote{\texttt{yaqd-brooks-mfc-gf} source code available on GitHub at \texttt{https://github.com/yaq-project/yaqd-brooks/}}, have been implemented in fewer than 100 lines of Python code.
Even the most complicated daemons are implemented in about 300 lines.
Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{pyserial, pyusb, pyvisa, pymodbus}.
In our experience, the process of creating a new daemon involves about a day of work after mastering communication with the hardware.
There are currently 72 daemons in the yaq project supporting at least 66 types of hardware, noting that some daemons support the same hardware and others are software only.
Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
A living list of all daemons and supported hardware can be found on the \yaq{} website.\footnote{See https://yaq.fyi/daemons/ and https://yaq.fyi/hardware/ for living lists of all \yaq{} daemons and supported hardware, respectively.}

@@

\yaq{} is open source software.
Anyone can view, install, edit, and suggest changes to our growing collection of daemons and clients.
Furthermore, anyone can create their own totally-custom client or daemon software separately by following the specified \yaq{} standard.
Thirteen individuals have contributed code to the development of \yaq{}.
Open-source development can be a powerful approach for research communities looking to share software-development and maintenance burdens. \cite{CohenJeremy2021a}
It is our hope that a vibrant open-source community will form around \yaq{}.
While open-source development is not a panacea \cite{NowogrodzkiAnna2019a}, we hope that by maintaining a distributed development strategy with a strict focus on only hardware interfaces, the \yaq{} project might prove sustainable.
